Query Optimization for Dynamic Imputation

### 2017-03-01 paper/algorithm.tex
paper: phillip changes to algorithm section

\section{Algorithm}
In order to correctly plan around missing values, we first extend the relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
with two new operators (\Cref{sec:operators}). We then define the search space (\Cref{sec:search-space}) of plans to encompass non-trivial operations, such as joins and 
aggregations. Soundness of query execution is provided by our main design invariant, which guarantees traditional relational
algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). 
The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates for the 
quality of the imputations performed and the runtime performance. Our imputation planning algorithm is
agnostic to the type of imputation used (\Cref{sec:imputation}). Finally, we show
that while our algorithm is exponential in the number of joins(\Cref{sec:complexity}), in practice, planning times are not a concern.

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all \nullv{} values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
In our experiments, we instantiate \textit{Impute} with the CE-CART~\cite{burgette2010multiple} algorithm, but we highlight that
our system is agnostic to this choice.
\textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no \nullv{} values for
attributes in $C$.

@@

\item Joins are performed after filtering, and only left-deep plans are considered.

@@

\subsection{Imputation placement}
\label{sec:placement}
We place imputation operators so that no relational operator encounters a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.

Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a \nullv{} value, there are cases where an imputation operator is required. To track these cases, each query plan $q$ is associated with a set of dirty attributes $\textsc{Dirty}(q)$. An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain \nullv{}. We compute a dirty set for each table using the table histograms, which track the number of \nullv{} values in each column. The dirty set for a query plan can be computed recursively as follows:

@@

\begin{definition}
We say that a plan $p$ with cost $\langle l, t \rangle$ \emph{dominates} a plan $p'$ with cost $\langle l', t' \rangle$ if $\textsc{Dirty}(p) = \textsc{Dirty}(p') \land ((l \leq l' \land t < t') \lor (l < l' \land t \leq t'))$. We denote this as $\doms{p}{p'}$. 
\end{definition}

@@

The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of a query plan $q$ and its sub-plans. 
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan, for example
redistributing the count of tuples with \nullv{} in an imputed field.
\Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

We use these histogram estimates to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.

@@

Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

To compute $\textsc{Time}(Q)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard 
database.

@@

\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal
effort and without changes to the optimizer, simply by adding time and loss functions
(\Cref{sec:cost-model}) which describe the new algorithm. In principle, new imputation algorithms
can be added, possibly targeted to a specific domain. The flexibility of this approach
aligns well with the spirit of query optimization, which aims to reconcile performance with
a declarative interface to data manipulation.

>>>>

\section{Algorithm}
In order to correctly plan around missing values, we first extend the relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
with two new operators (\Cref{sec:operators}). We then define the search space (\Cref{sec:search-space}) of plans to encompass non-trivial operations, such as joins and 
aggregations. Soundness of query execution is provided by our main design invariant, which guarantees traditional relational
algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). 
The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates for the 
quality of the imputations performed and the runtime performance. Our imputation planning algorithm is
agnostic to the type of imputation used (\Cref{sec:imputation}). Finally, we show
that while our algorithm is exponential in the number of joins (\Cref{sec:complexity}), in practice, planning times are not a concern.

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all \nullv{} values with non-\nullv{} values for attributes $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
In our experiments, we instantiate \textit{Impute} with the CE-CART~\cite{burgette2010multiple} algorithm, but we highlight that
our system is agnostic to this choice.
\textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no \nullv{} values for
attributes in $C$.

@@

\item Joins are performed after filtering, and only left-deep plans are considered.
  This significantly reduces the plan space while still allowing plans with interesting join patterns.

@@

\subsection{Imputation placement}
\label{sec:placement}
We place imputation operators into the query plan so that no relational operator encounters a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.

Imputation operators can be placed at any point, but there are cases where an imputation operator is required to meet the guarantee that no non-imputation operator sees a \nullv{} value. To track these cases, we associate each query plan $q$ with a set of dirty attributes $\textsc{Dirty}(q)$. An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain \nullv{}. We compute a dirty set for each table using the table histograms, which track the number of \nullv{} values in each column. The dirty set for a query plan can be computed recursively as follows:

@@

\subsubsection{Pareto-optimal plans}
\begin{definition}
We say that a plan $p$ with cost $\langle l, t \rangle$ \emph{dominates} a plan $p'$ with cost $\langle l', t' \rangle$ if $\textsc{Dirty}(p) = \textsc{Dirty}(p') \land ((l \leq l' \land t < t') \lor (l < l' \land t \leq t'))$. We denote this as $\doms{p}{p'}$. 
\end{definition}

@@

\subsubsection{Cardinality estimation}
The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of a query plan $q$ and its sub-plans.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.

To compute cardinality, we maintain histograms for each column in the database.
During query planning, each of the logical nodes in a query plan points to a set of histograms which describe the distribution of values in the output of the node.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan.
For example it will redistribute the count of tuples with \nullv{} in an imputed field.
\Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

\subsubsection{Imputation quality}
We use these histogram estimates to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.

@@

To add an alternate imputation strategy to \ProjectName{} could be used, the analyst must provide a corresponding loss with a reasonable parameterization.

\subsubsection{Query runtime}
To compute $\textsc{Time}(Q)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard 
database.

@@

\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal
effort and without changes to the optimizer, simply by adding time and loss functions
(\Cref{sec:cost-model}) which describe the new algorithm.
An imputation strategy is a function $I(T, C) \rightarrow T'$ which fills in the missing attributes $C$ in a set of tuples $T$, returning a new set of tuples $T'$.
In principle, new imputation algorithms
can be added, possibly targeted to a specific domain. The flexibility of this approach
aligns well with the spirit of query optimization, which aims to reconcile performance with
a declarative interface to data manipulation.
